CrypTool Script \\ Mathematics and Cryptography

### 2002-06-03 script/en/chapters/primes.tex
New script directory and file layout (only English for now)

\documentclass[a4paper,11pt]{article}
% Da man keine Unterst"utzung der deutschen Sprache in diesem Fall braucht!!
% \usepackage[german]{babel}

\setlength{\fboxrule}{.5mm}
\setlength{\fboxsep}{1.75mm}
\setlength{\footnotesep}{6pt}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{makeidx}
\usepackage[pdftex]{graphicx}
% \usepackage{graphicx}
\makeindex
\hypersetup{colorlinks=true}
\usepackage[flushmargin]{footmisc}

\setlength{\textwidth}{460 pt}
\hoffset-0.7in
\newtheorem{definition}{Definition}
%\newcommand{\href}[2]{#2}
\newcommand{\hypertag}[2]{#1}
%\newcommand{\hyperlink}[2]{#2}
%\newcommand{\hypertarget}[2]{#2}
\newtheorem{theorem}{Theorem}
\newenvironment{Proof}[1]
  {\textbf{Proof #1} \\}
  {\hfill$\Box$ \\}

%\pdfinfo{ /Title (CrypTool script)
%/Creator (TeX)
%/Producer (pdfTeX 0.14a)
%/Author (Deutsch Bank AG)
%/CreationDate (D:20000920201000)
%}

\title{ CrypTool Script \\ Mathematics and Cryptography }

\author{
(c) The authors, 1998-2002 \\
Frankfurt am Main
}

\begin{document}
\pagestyle{plain}
\setlength{\fboxrule}{.5mm}
\setlength{\fboxsep}{1.75mm}
\setlength{\footnotesep}{6pt}
\addtolength{\footskip}{8pt}
%\setlength{\footskip}{4cm}
%\renewcommand{\footnoterule}{\parindent0cm\rule{13cm}{.1pt}\vspace{.2cm}}
%Formatierung der Fu"snoten

%space between text and footnote 
\renewcommand\footnoterule{%
  \vspace{2em}%   <-- one line space between text and footnoterule
%\kern-3\p@
  \hrule width .4\columnwidth
 \vspace{4pt}
%kern 2.6\p@
}

%\long\def\@makefntext#1{%
%    \parindent 1em%
 %   \noindent
%    \hbox to 1.8em{\hss\@makefnmark}#1}

\maketitle

\parskip 4pt
\vskip + 30 pt
{
In this CrypTool script you will find predominantly mathematically oriented
information for using encryption procedures. The main chapters have been written
by {\em various authors} and are therefore independent from one another. At the end of
most chapters you will find literature and URLs.

You will receive information about the principles of symmetrical and
asymmetrical encryption. A large section of the script is dedicated to the
fascinating topic of prime numbers. Using numerous examples, the elementary number
theory and modular arithmetic are introduced and applied in an exemplary 
manner in the RSA procedure.  After this, one receives insight into the mathematical ideas
behind modern cryptography.

%You will also obtain an overview of the
%mathematical ideas behind modern cryptography.

A further chapter is devoted to digital signatures, which are an essential
component of e-business applications. The last chapter --- elliptic curves ---
fits in well with this. The mathematics of elliptic curves forms the basis for
rapid cryptographic algorithms for digital signatures; these algorithms are
extremely well suited for implementation on smartcards.

Whereas the program CrypTool teaches you how to use cryptography in practice, the
script provides those interested in the subject with a deeper understanding of
the mathematical algorithms used -- trying to do it didactically as good as possible.

The {\em authors} Bernhard Esslinger, Bartol Filipovic, Henrik Koy and Roger Oyono
would like to take this opportunity to thank their colleagues in the company and
at the universities of Frankfurt, Giessen and Karlsruhe. They particularly thank
Dr. Peer Wichmann from the Karlsruhe computer science research centre
(Forschungszentrum Informatik, FZI) for his down-to-earth support.

\enlargethispage{0.5cm}
As at CrypTool, the quality of the script is enhanced through your suggestions
and ideas for improvement. We look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool under \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},~~
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com}~~ or~~
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
{
\vskip + 7 pt \noindent
The contact persons for this free tool are listed in the readme file for
CrypTool.}
}

\newpage
\tableofcontents
\addcontentsline{toc}{section}{Contents}
\newpage
\normalsize
\parindent0cm
\parskip4pt

\section*{Introduction}  \addcontentsline{toc}{section}{Introduction}

This script is delivered together with CrypTool.

CrypTool is a program with an extremely comprehensive online help enabling you
to use and analyse cryptographic procedures.\par \vskip + 3pt

For the conventional encryption procedures, automatic analyses are available in CrypTool
that enable you to determine the key with knowledge of the encrypted document
and, possibly, further information (non-encrypted document or language of the
document).\par \vskip + 3pt

CrypTool was developed during the End-User Awareness program in order to
increase employee awareness of IT security and provide them with a deeper
understanding of the term security.\par \vskip + 3pt

A further aim was to enable users to understand the cryptographic procedures
implemented in the Deutsche Bank. In this way, using CrypTool as a reliable
reference implementation of the various encryption procedures (because of using the industry-proven SECUDE Library),
you can test the encryption implemented in other programs. \par \vskip + 3pt

Because the articles in this script are largely self-contained, this script can also
be read independantly of CrypTool.

The {\em authors} tried to describe cryptography for a broad audience --
without being mathematically incorrect. We believe, that this didactical pretention
is the way to promote the awareness for IT security and the readiness to use standardised modern cryptography.

\newpage
\section{Encryption procedures}

\subsection{Encryption}

The purpose of encryption \index{Encryption} is to change data in such a way
that only an authorised recipient is able to reconstruct the plaintext. This has
the advantage that you can transmit encrypted data openly and nevertheless need
not fear a perpetrator reading the data without authorisation. Authorised
recipients possess a piece of secret information --- called the key --- which
allows them to decrypt the data while it remains hidden from everyone else.\par \vskip + 3pt

One encryption procedure has been proved to be secure --- the {\em One-Time-Pad}.
\index{One­Time­Pad} However, this procedure has several practical
disadvantages (the key used must be selected randomly and must be just as long
as the message to be protected), which means that it is hardly used except in
closed environments such as for the hot wire between Moscow and Washington.\par \vskip + 3pt

For all other procedures there is a (theoretical) possibility of breaking them.
If the procedures are good, however, the time taken to break them is so long
that it is practically impossible to do and these procedures can therefore be
considered (practically) secure.\par \vskip + 3pt

We basically distinguish between symmetric and asymmetric encryption procedures.

\subsubsection{Symmetric encryption}

For {\em symmetric} encryption \index{Encryption!symmetric} the sender and
recipient must possess a common (secret) key which they have exchanged before
actually starting to communicate. The sender uses this key to encrypt the
message and the recipient uses it to decrypt it.\par \vskip + 3pt

The advantages of symmetric algorithms are the high speed with which data can be
encrypted and decrypted. One disadvantage is the need for key management. In
order to communicate with one another confidentially, sender and recipient must
have exchanged a key using a secure channel before actually starting to
communicate. Spontaneous communication between individuals who have never met
therefore seems virtually impossible. If everyone wants to communicate with
everyone else spontaneously at any time in a network of $ n $ subscribers, each
subscriber must have previously exchanged a key with each of the other $n-­ 1$
subscribers. A total of $n(n - 1)/2$ keys must therefore be exchanged.\par \vskip + 3pt

The most well-known symmetric encryption procedure is the \index{DES} DES--
algorithm. The DES-algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing all possible keys until the right one is found ({\em brute--force--attack}).
\index{Attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can now only be considered to be conditionally secure. Symmetric
alternatives to the DES procedure include the IDEA \index{IDEA} or Triple DES
algorithms.\par \vskip + 3pt

Up-to-the-minute procedures are the symmetric AES procedures. The associated
Rijndael procedure was declared winner the AES award on 2 October 2000 and thus
succeeds the DES procedure.

\subsubsection{Asymmetric encryption}

In the case of {\em asymmetric} encryption \index{Encryption!asymmetric} each
subscriber has a personal pair of keys consisting of a {\em secret}
\index{Key!secret} key and a {\em public} key\index{Key!public}. The public
key, as its name implies, is made public, e.g. in a key directory on the
Internet.\par \vskip + 3pt

If Alice wants to communicate with Bob, then she finds Bob's public key in the
directory and uses it to encrypt her message to him. She then sends this
ciphertext to Bob, who is then able to decrypt it again using his secret key. As
only Bob knows his secret key, only he can decrypt messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

Such a procedure can be demonstrated using a series of thief-proof letter boxes.
If I have composed a message, I then look for the letter box of the recipient
and post the letter through it. After that, I can no longer read or change the
message myself, because only the legitimate recipient possesses the key for the
letter box.\par \vskip + 3pt

The advantage of asymmetric procedures is the easy \index{Key
management} key management. Let's look again at a network with $n$
subscribers. In order to ensure that each subscriber can establish
an encrypted connection to each other subscriber, each subscriber
must possess a pair of keys. We therefore need $2n$ keys or $n$
pairs of keys. Furthermore, no secure channel is needed before
messages are transmitted, because all the information required in
order to communicate confidentially can be transmitted openly. In
this case, you simply have to pay attention to the accuracy
(integrity and authenticity) \index{Authenticity} of the public
key. Disadvantage: Pure asymmetric procedures take a lot longer to
perform than symmetric ones.\par \vskip + 3pt

The most well-known asymmetric procedure is the \index{RSA} RSA algorithm,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signatures!digital}Signature
Algorithm).

\newpage
\subsubsection{Hybrid procedures}
\index{Hybrid procedures}
In order to benefit from the advantages of symmetric and asymmetric techniques
together, hybrid procedures are usually used (for encryption) in practice.\par \vskip + 3pt

In this case the data is encrypted using symmetric procedures: the key is a
session key generated by the sender randomly that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management \index{Key
management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

\subsubsection{Further details}

Beside information you can find in many books and on a lot of websites the online help
of CrypTool also offers very many details about the symmetric and asymmetric encryption methods.

\begin{thebibliography}{99999}
\addcontentsline{toc}{subsection}{Literatur}
\bibitem[Schmeh2001]{Schmeh2001}  \index{Schmeh 2001}
	Klaus Schmeh, \\
	Kryptografie und Public-Key Infrastrukturen im Internet, dpunkt.verlag, 2. akt. und erw. Auflage 2001. \\
	A considerable, up-to-date, good reading book, which considers practical problems, like standardisation or
	real existing software. Currently published only in German language.
\end{thebibliography}

@@

Every whole number has a factor. The number 1 only has one factor,
itself, whereas the number $12$ has the six factors $1, 2, 3, 4,
6, 12$. Many numbers can only be divided by themselves and by $1$.
With respect to multiplication, these are the ''atoms'' in the
area of numbers. Such numbers are called prime numbers.

@@

All whole numbers greater than $1$ can be expressed as a product of prime
numbers --- in a unique way. This is the claim of the 1st fundamental theorem of
number theory (= fundamental theorem of arithmetic = fundamental building block
of all positive integers).

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
egyptian academy of Alexandria and wrote ''The Elements'', the most well known systematical textbook
of the Greek mathematics.} knew this in the
third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as following
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

The largest prime numbers known today have several hundred
thousand digits, which is too big for us to imagine. The number of
elementary particles in the universe is ''only'' estimated to be a
$80$-digit number \hyperlink{grosord}{(See: overview under 2.6.3
of various orders of magnitude / dimensions)}.

@@

\index{Numbers!Mersenne numbers}\index{Mersenne Marin!Mersenne numbers}
\index{Mersenne Marin!theorem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). A prime Mersenne number \index{Mersenne prime number} is called Mersenne prime number.  
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

Due to the fact that the exponents of the Mersenne numbers do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers \index{Mersenne prime number} 
have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253; ~ 4,423; \\
9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; ~ 216,091;\\
 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; ~13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known.

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne Marin!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne Marin!M-39}

@@

Now more than 130,000 volunteers, amateurs and experts, are working for the GIMPS project. They connect their
computers into the so called ''primenet'', organized by the company entropia, to find such numbers using distributed
computer programs.

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[''small'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime number}\index{Numbers!pseudoprime
number} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Numbers!Carmichael numbers} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

    This sequence is also defined recursively and increases extremely quickly.
Does it only consist of primes?
$$
\begin{array}{lll}
C(0) = 2 & & \mapsto {\rm ~prime}\\
C(1) = 2^2 - 1 &    = 3 & \mapsto {\rm ~prime}\\
C(2) = 2^3 - 1 &    = 7 & \mapsto {\rm ~prime} \\
C(3) = 2^7 - 1 &    = 127& \mapsto {\rm ~prime} \\
C(4) = 2^{127} - 1 &      = 170, 141, 183, 460, 469, 231, 731, 687, 303, 715,
884, 105, 727 & \mapsto {\rm ~prime} \\
\end{array} 
$$
It is not (yet) known whether $C_5$ and higher elements are prime, but this is
not very likely.
In any case, it has not been proved that this formula delivers only primes.
\end{enumerate}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.e. valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richtstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ''Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250 old puzzle. To
stimulate the sell rates the English and American publisher compete a prize of 1 million USD, 
is someone proves the conjecture -- to be published till 2004 by an well-known mathematical journal.\\
Susprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ''Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

\bibitem[Schneier1996]{Schneier1996} \index{Schneier 1996}     Bruce Schneier,
\\     Applied Cryptography, Wiley and Sons, 2nd edition, 1996.

@@

\input{elementaryNumberTheory.inc}

\input{modernCryptology.inc}

\newpage
\section{Digital signatures}
\index{Signatures!digital}
The aim of digital signatures is to guarantee the following two points:
\begin{itemize}
 \item User authenticity: \\
      \index{Authenticity!user} It can be checked whether a message really does
come from a particular person.
 \item Message integrity: \\
      \index{Message integrity} It can be checked whether the message has been
changed (on route).
\end{itemize}

An asymmetric technique is used again (see encryption procedures). Participants
who wish to generate a digital signature for a document must possess a pair of
keys. They use their secret key to generate signatures and the recipient uses
the sender's public key to verify whether the signature is correct. As before,
it must be impossible to use the public key to derive the secret key.

In detail, a \index{Signature procedure} {\em Signature procedure} looks like
this: \\ Senders use their message and secret key to calculate the digital
signature for the message. Compared to hand-written signatures, therefore,
digital signatures have the advantage that they also depend on the document to
be signed. Signatures from one and the same participant are different unless the
signed documents are completely identical. Even inserting a blank in the text
would lead to a different signature. The recipient of the message would
therefore detect any injury to the message integrity as this would mean that the
signature no longer matches the document and is shown to be incorrect when
verified.

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. In practice, however, the
procedure we have just described has a decisive disadvantage. The signature is
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance, we use a cryptographic
hash function.

>>>>

@@

Every whole number has a factor. The number 1 only has one factor,
itself, whereas the number $12$ has the six factors $1, 2, 3, 4,
6, 12$. Many numbers can only be divided by themselves and by $1$.
With respect to multiplication, these are the ``atoms'' in the
area of numbers. Such numbers are called prime numbers.

@@

All whole numbers greater than $1$ can be expressed as a product of prime
numbers --- in a unique way. This is the claim of the 1st fundamental theorem of
number theory (= fundamental theorem of arithmetic = fundamental building block
of all positive integers).\index{Number theory!fundamental theorem}

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
egyptian academy of Alexandria and wrote ``The Elements'', the most well known systematical textbook
of the Greek mathematics.} knew this in the
third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as following
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

The largest prime numbers known today have several hundred
thousand digits, which is too big for us to imagine. The number of
elementary particles in the universe is ``only'' estimated to be a
$80$-digit number \hyperlink{grosord}{(See: overview under 2.6.3
of various orders of magnitude / dimensions)}.

@@

\index{Numbers!Mersenne}\index{Mersenne!number}
\index{Mersenne!theorem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). A prime Mersenne number \index{Mersenne!prime number} is called Mersenne prime number.  
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers \index{Mersenne!prime number} 
have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known.

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne!number!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne!number!M-39}

@@

Now more than 130,000 volunteers, amateurs and experts, are working for the GIMPS project. They connect their
computers into the so called ``primenet'', organized by the company entropia, to find such numbers using distributed
computer programs.

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[``small'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime numbers}\index{Numbers!pseudoprime
number} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Numbers!Carmichael numbers} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

    This sequence is also defined recursively and increases extremely quickly.
Does it only consist of primes?
$$
\begin{array}{lll}
C(0) = 2 & & \mapsto {\rm \:prime}\\
C(1) = 2^2 - 1 &    = 3 & \mapsto {\rm \:prime}\\
C(2) = 2^3 - 1 &    = 7 & \mapsto {\rm \:prime} \\
C(3) = 2^7 - 1 &    = 127& \mapsto {\rm \:prime} \\
C(4) = 2^{127} - 1 &      = 170, 141, 183, 460, 469, 231, 731, 687, 303, 715,
884, 105, 727 & \mapsto {\rm \:prime} \\
\end{array} 
$$
It is not (yet) known whether $C_5$ and higher elements are prime, but this is
not very likely.
In any case, it has not been proved that this formula delivers only primes.
\end{enumerate}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.e. valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richtstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250 old puzzle. To
stimulate the sell rates the English and American publisher compete a prize of 1 million USD, 
is someone proves the conjecture -- to be published till 2004 by an well-known mathematical journal.\\
Susprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

%already defined in elementaryNumberTheory.inc
\item[Schneier1996]{Schneier1996} \index{Schneier 1996}     Bruce Schneier,
\\     Applied Cryptography, Wiley and Sons, 2nd edition, 1996.

@@
